On the entropy of isothermal dusty gas: application to protoplanetary disks

### 2016-08-04 model.tex
minor edits at home

\subsection{Locally isothermal equation of state for the gas} 
We adopt 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \tepsilon)\rho,   
\end{align}
where $c_s$ is a prescribed sound-speed profile fixed in time. 
In numerical calculations we will consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature.

If we define the effective sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
    \tepsilon\right)}$, then the equation of state has the standard
form $P=c_{s,\mathrm{eff}}^2\rho$. Since the dust-fraction $\tepsilon$
typically decrease away from the midplane, we expect vertically
isothermal dusty disks to in fact behave as if it
had a temperture increasing away from $z=0$.

@@

@@

Finally, we define 
\begin{align}
  Z \equiv \epsilon_0\frac{\Hd}{\Hg} \simeq
  \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}} 
\end{align}
as a measure of the local metalicity. The second approximation holds
for $\epsilon_0\ll1$. 
%\Sigma_\mathrm{d}$ and the gas
%surface density $\Sigma_\mathrm{g}$.

@@

\subsection{Vertical shear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right),
\end{align}
where
\begin{align}
   s \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]} 
\end{align}
is a pseudo-entropy for the mixture.

Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\epsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r}\left[\frac{\p\epsilon}{\p z} + \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

We can compare these sources by
evaluating them, to order-of-magnitude, using the equilibrium
solutions in \S\ref{steady_state}. We assume the disk is radially
smooth so that $\p_r\sim 1/r$, and the dust-to-gas ratio
$\epsilon\ll1$. This gives

@@

\subsection{Vertical buoyancy}
Having identified the entropy of an isothermal dusty gas, we find the
vertical buoyancy frequency $N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p s}{\p z} &=
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
  z}\frac{\p\epsilon}{\p z} \\ &
                                  =
  \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where the second equality assumes $c_s=c_s(r)$ and the final equality
applies to the equilibria in \S\ref{eqm}. We find 
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
Thus in PPDs where $\epsilon_0 = O(10^{-1}$---$10^{-2})$, 
vertical buoyancy only has a modestly stabilizing effect.

>>>>

\subsection{Locally isothermal equation of state} 
We adopt 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \tepsilon)\rho,   
\end{align}
where $c_s$ is a prescribed sound-speed profile fixed in time. 
In numerical calculations we will consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

If we define the effective sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
    \tepsilon\right)}$, then the equation of state has the standard
form $P=c_{s,\mathrm{eff}}^2\rho$. Since the dust-fraction $\tepsilon$
typically decrease away from the midplane, we expect vertically
isothermal dusty disks to in fact behave as if it
had a temperture \emph{increasing} away from $z=0$.

@@

\subsection{Entropy of an isothermal dusty gas }
%Having made the connection between the dust fraction (or dust-to-gas
%ratio) of the mixture and an energy equation, we now define 
We define the (dimensionless) entropy of the mixture as 
\begin{align}
   s \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]}.  
\end{align}
We shall see that by defining the entropy this way, many of the 
results concerning the stability of dusty gas will have identical form
and interpretations to that in standard adiabatic hydrodynamics.

@@

Finally, we define 
\begin{align}
  Z \equiv \epsilon_0\frac{\Hd}{\Hg} \simeq
  \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}} 
\end{align}
as a measure of the local metalicity, where $\Sigma_\mathrm{d}$ and
$\Sigma_\mathrm{g}$ are the dust and gas surface densities,
respectively. The second equality holds for $\epsilon_0\ll1$.

%\Sigma_\mathrm{d}$ and the gas
%surface density $\Sigma_\mathrm{g}$.

@@

\subsection{Vertical shear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\epsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r}\left[\frac{\p\epsilon}{\p z} + \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

We can compare these sources by
evaluating them using the equilibrium
solutions in \S\ref{steady_state}. We assume the disk is radially
smooth so that $\p_r\sim 1/r$, and the dust-to-gas ratio
$\epsilon\ll1$. This gives

@@

\subsection{Vertical buoyancy}
Having identified the entropy of an isothermal dusty gas, we find the
vertical buoyancy frequency $N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p s}{\p z} &=
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
  z}\frac{\p\epsilon}{\p z} \\ &
                                  =
  \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where the second equality assumes $c_s=c_s(r)$ and the final equality
applies to the equilibria in \S\ref{eqm}. Thus,  
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
However, for thin dust layers such that $H_\epsilon \ll Hg$, 
$\mathrm{max}\left(N_z\right)$ may occur outside a finite vertical domain.

We see that $N_z^2\sim \epsilon \OmK^2$, while $r\p_z\Omega^2\sim
h_\mathrm{g}\OmK^2$ for a thin disk where vertical shear is mainly due to the
radial temperature gradient. Since buoyancy is stabilizing, we expect
the dust-induced buoyancy forces to stabilize the disk against the VSI
where $\epsilon \gtrsim h_\mathrm{g}$.

### 2016-08-04 paper.tex
minor edits at home

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  %Tightly-coupled dust particles
  % in protoplanetary disks almost follow the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
\begin{inparaenum}[1)]
\item
  identify the entropy of isothermal dusty gas; and 
\item 
  generalize previous, well-known gaseous instabilities in
  protoplanetary disks to include dust dynamics. 
\end{inparaenum}
This includes the axisymmetric stability of strictly isothermal dusty
gas, allowing us to constrain the sharpness of 
dust edges that can persist in protoplanetary disks. We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results are consistent with our definition of
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss potential applications of the 
equivalence between isothermal dusty gas and standard hydrodynamics
in providing thermodynamical interpretations of other dusty
instabilities in protoplanetary disks.  
\end{abstract}

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  %Tightly-coupled dust particles
  % in protoplanetary disks almost follow the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
\begin{inparaenum}[1)]
\item
  identify the entropy of isothermal dusty gas; and 
\item 
  generalize previous, well-known gaseous instabilities in
  protoplanetary disks to include dust dynamics. 
\end{inparaenum}
This includes the axisymmetric stability of strictly isothermal dusty
gas, allowing us to constrain the sharpness of 
dust edges that can persist in protoplanetary disks. We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results are consistent with our definition of
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss potential applications of the dusty/adiabatic gas
correspondence to provide thermodynamical interpretations of dusty
instabilities in protoplanetary disks.  
\end{abstract}

### 2016-08-04 summary.tex
minor edits at home

\section{Summary and discussion}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. 
In the limit of perfect dust-gas coupling, the dust content of a gas 
parcel $\rhod/\rhog$ is conserved. This is analogous to entropy being 
conserved following a adiabatic, pure gas. %s = s(\rhod/\rhog)
We explicitly show that for 
a locally isothermal, dusty gas that the evolutionary equation of 
$\rhod/\rhog$ may be recast into an effective energy equation identical
in form to that in standard adiabatic hydrodynamics. We find the 
natural definition of the entropy of the dusty gas is 
\begin{align*}
  s  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
This allows us to define the buoyancy frequency of an isothermal 
dusty gas.

@@

We obtain the Solberg-Hoiland criteria to assess
axisymmetric stability when the equation of state is strictly
isothermal. We show that {\bf dust-settling cannot lead to
  axisymmetric instability, however large the associated vertical
  shear} \citep[cf. \emph{non-axisymmetric} Kelvin-Helmholtz instabilities
  induced by  dust-settling, ][]{lee10}. We  also show that
protoplanetary disks cannot have sharp edges in the dust-to-gas ratio,
as this implies a sharp entropy gradient and may thus render the disk
unstable.

@@

For locally isothermal disks $c_s=c_s(r)$ we generalize the vertical
shear instability (VSI) to include dust. We find that dust-loading
generally has a stabilizing effect. In our disk models
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed at heights where vertical 
buoyancy dominates over vertical shear. This confirms our 
identification of the entropy and buoyancy of an isothermal, dusty
gas.

@@

\subsection{Dusty analogs of other gaseous instabilities}

%gi
%rwi

>>>>

\section{Summary and discussion}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. 
In the limit of perfect dust-gas coupling, the dust content of a fluid 
parcel $\rhod/\rhog$ is conserved. This is analogous to entropy being 
conserved following a adiabatic, pure gas. %s = s(\rhod/\rhog)
We explicitly show that for 
a locally isothermal, dusty gas that the evolutionary equation of 
$\rhod/\rhog$ may be recast into an effective energy equation identical
in form to that in standard adiabatic hydrodynamics. We find the 
natural definition of the entropy of the dusty gas is 
\begin{align*}
  s  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
This allows us to define the buoyancy frequency of an isothermal 
dusty gas.

@@

We obtain the Solberg-Hoiland criteria to assess
axisymmetric stability when the equation of state is strictly
isothermal. We show that {\bf dust-settling cannot lead to
  axisymmetric instability, however large the associated vertical
  shear} \citep[cf. \emph{non-axisymmetric} Kelvin-Helmholtz instabilities
  induced by  dust-settling, ][]{lee10}. We also show that
sharp edges in the dust-to-gas ratio cannot persist in protoplanetary
disks, as these imply a sharp entropy gradient and may thus destablize
the disk.

@@

For locally isothermal disks $c^2_s=c_s^2(r)$ we generalize the vertical
shear instability (VSI) to include dust. We find that dust-loading
generally has a stabilizing effect. In our disk models
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed at heights where vertical 
buoyancy dominates over vertical shear. This confirms our 
identification of the entropy and buoyancy of an isothermal, dusty
gas.

@@

\subsection{Dusty analogs of other gaseous instabilities}  
%gi
%rwi
